---
title: "Assignment 1 - Language Development in ASD - part 4"
author: "Amalie Lysgaard Andersen"
date: "August 10, 2017"
output: html_document
---
```{r}
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/4. Aarhus Universitet/3. Semester/1. Experimental Methods lll/R FOLDER/languagedevelopmentASD/languagedevelopmentASD")

library(pacman)
p_load(lmerTest, pastecs, ggplot2, tidyverse, gdata, MuMIn, effects, stringr, plyr, Metrics, vtreat, cvTools, simr, lme4)
update.packages("lmer")
```

## Welcome to the fourth exciting part of the Language Development in ASD exercise

In this exercise we will assess how many participants we would need to adequately replicate our findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8).

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- [GitHub]Load your dataset, fit your favorite model, assess power for your main effects and interactions of interest.
- Report the power analysis and comment on what you can (or cannot) use its estimates for. (giver det mening at lave dette eksperiment med det data, og variable vi har)
```{r Load + clean data, include = F}
# Load data  
data = read.csv("LanguageASD.csv")  

# Changing names in the training data
data <- rename.vars(data, "Diagnosis", "diagnosis") 
data <- rename.vars(data, "Gender", "gender")
data <- rename.vars(data, "Age", "age")
data <- rename.vars(data, "Ethnicity", "ethnicity")
data <- rename.vars(data, "Visit", "visit")
data <- rename.vars(data, "ADOS1", "ados1")
data <- rename.vars(data, "MullenRaw1", "nonverbalIQ")
data <- rename.vars(data, "ExpressiveLangRaw1", "verbalIQ")

# Filtering out participants 
data <- data[-c (1, 68, 69, 130, 131, 132, 276, 277, 284, 285), ] 

# Removing NA's 
data <- na.omit(data)
 
# Changing to numeric
data$visit <- as.numeric(data$visit)
data$ID <- as.numeric(data$ID)

str(data)
```

```{r Model + Power for effects, include = F}
# Models 
# zeus <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data, REML=FALSE)

# artemis <-  lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + diagnosis + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data, REML=FALSE)

# aphrodite <-  lmer(CHI_MLU ~ visit * diagnosis + visit + I(visit^2) + I(visit^3) + diagnosis + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data, REML=FALSE)
#summary(aphrodite)

# Best model from the previous assignments
bestmodel <- lmer(CHI_MLU ~ visit*diagnosis + I(visit^2) + verbalIQ + (1 + visit + I(visit^2) | ID), data, REML = F)
summary(bestmodel)






# Calculating power for visit
powerV <- powerSim(bestmodel, simr::fixed('visit', method = 't'), nsim = 50)

 # Calculating power for quadratic visit
powerV2 <- powerSim(bestmodel, simr::fixed('I(visit^2)', method = 't'), nsim = 50)

# Calculating power for ASD severity
powerA = powerSim(bestmodel, simr::fixed("ados1", method = "t"),nsim=200)

# Calculating power for mother's MLU
powerM <- powerSim(bestmodel, simr::fixed('MOT_MLU', method = 't'), nsim = 200)

# Calculating power for verbal IQ
powerIQ <- powerSim(bestmodel, simr::fixed('verbalIQ', method = 't'), nsim = 200)

# Calculating power for diagnosis
powerD <- powerSim(bestmodel, simr::fixed('diagnosisTD', method = 't'), nsim = 1000)

# Calculating power for interaction: visit+diagnosis
powerVD <- powerSim(bestmodel, simr::fixed('visit:diagnosisTD', method = 't'), nsim = 1000)

lastResult()$warnings
# Getting the powers
powerIQ
powerD
powerV
powerV2
powerM
powerVD
powerA

## Testing
#fixef(bestmodel)["visit"]
#fixef(bestmodel)["ados1"]
#fixef(bestmodel)["diagnosisTD"]
#fixef(bestmodel)["verbalIQ"]
#fixef(bestmodel)["visit:diagnosisTD"]
```
# ANSWER 
For this exercise we chose our 'bestmodel' (a bit revised from previous assignments). We chose a quadratic linear mixed effects model predicting child MLU development from number of visit, diagnosis, mother's mean length of utternace and verbalIQ. Moreover, we included visit and diagnosis as an interaction. The model also takes ID as random intercept and random slopes by visit (model: Child mean length of utterance ~ visit * diagnosis + visit^2 + MOT_MLU + verbalIQ + (1 + visit + visit^2 | ID).

Power simulations were run on the relevant fixed effects and interaction effect of diagnosis and visits. Assuming that the model has the right estimates, the model's power analysis has the following result:

For the main effects: visit, mot_mlu, diagnosis and verbalIQ 

- PowerVisit =      99.00% (96.43, 99.88)    Effect size for visit is 0.32
- PowerVisit^2 =    94.50% (90.37, 97.22)    Effect size for I(visit^2) is -0.036
- PowerDiagnosis =  7.90% ( 6.30, 9.75)      Effect size for diagnosisTD is 0.050
- PowerVerbalIQ =   100.0% (98.17, 100.0)    Effect size for verbalIQ is 0.059
- PowerMot_mlu =    100.0% (98.17, 100.0)    Effect size for MOT_MLU is 0.29
- Interaction =     100.0% (99.63, 100.0)    Effect size for visit:diagnosisTD is 0.25

Given the 80% power threshold, the power calculated here indicates that almost all the variables have enough power to be used in this study. Diagnosis by itself is quite low, however it makes no sense to look this predictor alone, since an interaction effect was found. The power analysis indicates that the study was somewhat overpowered - in that many predictors have a power of nearly 100% which is beyond threshold. Nevertheless, this conclusion cannot be drawn solely based on this power analysis, since it is based on the estimates that our 'best model' predicted. These might be overestimates of the true effect size.



Walkthrough in class:
- When specifying the effect you're testing, Malte suggests adding 
  E.g. fixed("visit", method = "t")  
- Remember to remove NA's
- Be critical about 100 % effect sizes all the time. 





### Exercise 2
How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- [GitHub] take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
```{r Minimum effect size}
# Estimates = fixed effects
fixef(bestmodel)["visit"] <- 0.3
fixef(bestmodel)["diagnosisTD"] <- 0.05
fixef(bestmodel)["MOT_MLU"] <- 0.3
fixef(bestmodel)["verbalIQ"]<- 0.01
fixef(bestmodel)["visit:diagnosisTD"]<- 0.1

summary(bestmodel)

powerIQf <- powerSim(bestmodel, simr::fixed("visit:diagnosisTD", method = 't'), nsim = 50)
powerIQf

# Getting a power curve
pc<-powerCurve(bestmodel, simr::fixed("visit:diagnosisTD", method='t'), along = 'ID', nsim = 50)
pc
plot(pc)
```

- [GitHub] assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
```{r}
# Assessing power curve 
pc<-powerCurve(model, simr::fixed('verbalIQ', method='t'), along = 'ID', nsim = 50)
pc
plot(pc)

pc2 <-powerCurve(model, simr::fixed('visit:diagnosisTD', method='t'), along = 'ID', nsim = 50)
pc2
plot(pc2)

pc3 <-powerCurve(model, simr::fixed('MOT_MLU', method='t'), along = 'ID', nsim = 50)
pc3
plot(pc3)

pc4 <-powerCurve(bestmodel, simr::fixed('visit', method='t'), along = 'ID', nsim = 50)
pc4
plot(pc4)

pc5 <-powerCurve(model, simr::fixed('I(visit^2)', method='t'), along = 'ID', nsim = 50)
pc5
plot(pc5)

pc6 <-powerCurve(model, simr::fixed('diagnosisTD', method='t'), along = 'ID', nsim = 50)
pc6
plot(pc6)
```

- [GitHub] if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.




### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why
```{r}


set.seed(30)

#list with asd kids
ASD <- data %>%
  filter(diagnosis == "ASD")
  
#list with TD kids       
TD <- data %>%
  filter(diagnosis == "TD")

#15 random IDs in asd-list
asd15 <- sample(unique(ASD$ID), 15)
asd15 <- as.vector(asd15)

#15 random IDs in td-list
td15 <- sample(unique(TD$ID), 15)
td15 <- as.vector(td15)

#list of the 30 rndom selected
all30 <- c(asd15, td15)

#make dataframe with the chosen kids
data30 <- data %>%
  subset(ID %in% all30)

# Now doing the power analysis
set.seed(44)
# Power analysis
model30 <- lmer(CHI_MLU ~ visit*diagnosis + I(visit^2) + MOT_MLU + verbalIQ + (1 + visit + I(visit^2) | ID), data30, REML = F)
summary(model30)

```







#-------------------------------------------------------------------------------------------------------------------------


## Welcome to the fourth exciting part of the Language Development in ASD exercise

In this exercise we will assess how many participants we would need to adequately replicate our findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8).

### Exercise 1

How much power does your study have (if your model estimates are quite right)?

- [GitHub]Load your dataset, fit your favorite model, assess power for your main effects and interactions of interest.

```{r + Model}
#Best model from the previous assignments
bestmodel <- lmer(CHI_MLU ~ diagnosis + visit + I(visit^2) + verbalIQ + MOT_MLU + (1 + visit + I(visit^2) | ID), data = data)
summary(bestmodel)

#Changing to interaction model
ritter <- lmer(CHI_MLU ~ diagnosis*(visit + I(visit^2)) + verbalIQ + (1+ visit + I(visit^2) | ID), data = data, REML = F)
summary(ritter)

#Plotting the interaction - to see why visit^2 is not significant
fe <- summary(ritter)$coefficients[1:6,1]
fe

# Plot
plot_df <- expand.grid(visit=c(1,2,3,4,5,6),diagnosis=c(0,1)) %>% 
  mutate(CHI_MLU = fe[1] + fe[2]*diagnosis + fe[3]*visit + fe[4]*visit*visit + fe[5]*diagnosis*visit + fe[6]*diagnosis*visit*visit)

plot_interaction <- ggplot(plot_df, aes(y=CHI_MLU, x = visit, color = as.factor(diagnosis))) + 
  geom_line(size=2) + 
    labs(color = "diagnosis") + 
    scale_color_brewer(palette = "Dark2") + 
  theme(legend.position = "top") + 
  theme_classic()

plot_interaction
#seems very close to linear, esp. ASD
#Linear model w. interaction
linedanser <- lmer(CHI_MLU ~ diagnosis*visit + verbalIQ + (visit | ID), data = data, REML = F)
summary(linedanser)
```

NOTE: Now with interaction and thereby including diagnosis

```{r Power calculation}
#Power simulation for diagnosis
powerD <- powerSim(linedanser, simr::fixed('diagnosisTD', method = 't'), nsim = 200)
#Power simulation for visit
powerv <- powerSim(linedanser, simr::fixed('visit', method = 't'), nsim = 200)
#Power simulation for interaction
powerVD <- powerSim(linedanser, simr::fixed('diagnosisTD:visit', method = 't'), nsim = 200)
#Power simulation for verbalIQ
powerIQ <- powerSim(linedanser, simr::fixed('verbalIQ', method = 't'), nsim = 200)

powerD
#powerv
#powerv2
powerVD
powerIQ

```

- Report the power analysis and comment on what you can (or cannot) use its estimates for.

For this exercise we chose our 'bestmodel' (a bit revised from previous assignments). We chose a quadratic linear mixed effects model predicting child MLU development from number of visit, diagnosis, mother's mean length of utternace and verbalIQ. Moreover, we included visit and diagnosis as an interaction.The model also takes ID as random intercept and random slopes by visit (pseudo-code: Child mean lenght of utterance ~ visit * diagnosis + visit^2 + MOT_MLU + verbalIQ + (1 + visit + visit^2 | ID). 

Power simulations were run on the relevant fixed effects and interaction effect of diagnosis and visits. Assuming that the model has the right estimates, the model's power analysis has the following result:


For the main effects: visit, diagnosis and verbalIQ

Visit
97  %     (93.58, 98.89)
0.10

Diagnosis
86  %     (80.41, 90.49)
-0.42

Verbal IQ
100 %     (98.17, 100.0)
0.075

Interaction (Diagnosis x Visit)
100 %     (98.17, 100.0)
0.25


The power analysis indicates that the study was somewhat either overpowered in the sense that the predictors have a power of nearly 100% which is beyond the threshold. 
Nevertheless, this conclusion cannot be drawn solely based on this power analysis, since it is based on the estimates that our model predicted. These might be overestimates of the true effect size. 

The estimates are interesting in several ways. They can be used a priori to uncover how many participants are needed to ensure meaningful findings. Or - like here - they can be used afterwards to investigate the effect size possible with the number of participants the study is based on.


### Exercise 2

How would you perform a more conservative power analysis?

- Identify and justify a minimum effect size for each of your relevant effects

- [GitHub] take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.

```{r Resetting estimates}
#estimates = fixed effects
#Irrelevant due to significant interaction effect 
fixef(linedanser)["diagnosisTD:visit"]<- 0.15
fixef(linedanser)["verbalIQ"]<- 0.025

#Power simulations of set effect sizes
powerVDset <- powerSim(linedanser, simr::fixed('diagnosisTD:visit', method = 't'), nsim = 200)
powerVDset
powerIQset <- powerSim(linedanser, simr::fixed('verbalIQ', method = 't'), nsim = 200)
powerIQset
```

- [GitHub] assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect

```{r}
# assessing power curve for relevant effects
#VerbalIQ
pcIQ <-powerCurve(linedanser, simr::fixed('verbalIQ', method='t'), along = 'ID', nsim = 200)
pcIQ
plot(pcIQ)
#Interaction
pcIA <-powerCurve(linedanser, simr::fixed('diagnosisTD:visit', method='t'), along = 'ID', nsim = 200)
pcIA
plot(pcIA)
#Visit
pcV <-powerCurve(linedanser, simr::fixed('visit', method='t'), along = 'ID', nsim = 50)
pcV
plot(pcV)
#Diagnosis
pcD <-powerCurve(linedanser, simr::fixed('diagnosisTD', method='t'), along = 'ID', nsim = 50)
plot(pcD)
pcD
```

- [GitHub] if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis

- Report the power analysis and comment on what you can (or cannot) use its estimates for.

The estimate of the interaction was lowered in an attempt to make a more conservative power analysis. As our hypotheses refer to the effect of time and diagnosis on child MLU, the interaction effect represents our two main effects and thus remains of primary interest. We consider verbal IQ to be sufficiently conservative as is and have decided to keep it at 0.075. 

Using the chosen minimum effect sizes, these power curves (figures 2-4) indicate that 35 participants would be adequate to reach a threshold of 0.80 for the interaction effect. However, one might notice that if the had study been centred around the effect of verbal IQ, one could argue that ≈ 9 participants would be sufficient


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}
#Interaction 
pcIAbreak <-powerCurve(linedanser, simr::fixed('diagnosisTD:visit', method='t'), along = 'ID', nsim = 200, breaks = 25:40)
pcIAbreak
plot(pcIAbreak)
#VerbalIQ
pcIQbreak <-powerCurve(linedanser, simr::fixed('verbalIQ', method='t'), along = 'ID', nsim = 200, breaks = 25:40)
pcIQbreak
plot(pcIQbreak)
#Visit
pcVbreak <-powerCurve(linedanser, simr::fixed('visit', method='t'), along = 'ID', nsim = 50, breaks = 25:40)
pcVbreak
plot(pcVbreak)
#Diagnosis
pcDbreak <-powerCurve(linedanser, simr::fixed('diagnosisTD', method='t'), along = 'ID', nsim = 50, breaks = 25:40)
pcDbreak
plot(pcDbreak)
```





