---
title: "Assignment 1 - Language Development in ASD - part 3"
author: "Amalie Lysgaard Andersen"
date: "August 10, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("~/Library/Mobile Documents/com~apple~CloudDocs/4. Aarhus Universitet/3. Semester/1. Experimental Methods lll/R FOLDER/languagedevelopmentASD/languagedevelopmentASD")

library(pacman)
p_load(lmerTest, pastecs, ggplot2, tidyverse, gdata, MuMIn, effects, stringr, plyr, Metrics, vtreat, cvTools)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time, then learning how to cross-validate models and finally how to systematically compare models.

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:

* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=1
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=1
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=1

```{r Loading new data,  include=FALSE}
#Loading the new data 
demo_test <- read.csv("demo_test.csv")
token_test <- read.csv("token_test.csv")
lu_test <- read.csv("lu_test.csv")
```

```{r Cleaning new data,  include=FALSE}
#Renaming the data sets to have the same headline for the subject = ID 
demo_test <- rename.vars(demo_test, "Child.ID", "ID", info=T)
lu_test <- rename.vars(lu_test, "SUBJ", "ID", info=T)
token_test <- rename.vars(token_test, "SUBJ", "ID", info=T)
 
#Renaming the visit to have the same 
lu_test <- rename.vars(lu_test, "VISIT", "Visit", info=T)
token_test <- rename.vars(token_test, "VISIT", "Visit", info=T)

#Homogenise the way visit is reported, use of stringr to extract only numbers
lu_test$Visit <- str_extract(lu_test$Visit, "[1-6]")
token_test$Visit <- str_extract(lu_test$Visit, "[1-6]")
 
#Removing all punctuations in the ID column 
demo_test$ID <- str_replace_all(demo_test$ID, "[:punct:]", "")
lu_test$ID <- str_replace_all(lu_test$ID, "[:punct:]", "")
token_test$ID <- str_replace_all(token_test$ID, "[:punct:]", "")

#Selecting the variables needed
demo_test_sub <- select(demo_test, c(ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS,  MullenRaw, ExpressiveLangRaw))
lu_test_sub <- select(lu_test, c(ID, Visit, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd))
token_test_sub <- select(token_test, c(ID, Visit, types_MOT, types_CHI, tokens_MOT, tokens_CHI))

#Using join to merge to three datasets by ID and Visit
test <- join(demo_test_sub, lu_test_sub)
test <- join(test, token_test_sub)

#Ados for only visit 1
#Using subset to make a dataset only consisting of the data from visit 1
subset_test <- subset(test, Visit == 1, select = c(ID, ADOS, MullenRaw, ExpressiveLangRaw))

#Changing the names in the new data set 
subset_test <- rename.vars(subset_test, c("ADOS", "MullenRaw", "ExpressiveLangRaw"), c("ADOS1", "NonverbalIQ", "VerbalIQ"))

#Merging the two datasets using join 
test <- join(test, subset_test)

#Using select to make a clean dataset with variables in the right order
clean_test <- select(test, c(ID, Diagnosis, Visit, Gender, Age, Ethnicity,  ADOS1, NonverbalIQ, VerbalIQ, CHI_MLU, MOT_MLU, CHI_LUstd, MOT_LUstd, types_CHI, types_MOT,  tokens_CHI, tokens_MOT))

#Anonymising kids
clean_test$ID <- as.factor(clean_test$ID)
clean_test$ID <- as.numeric(clean_test$ID)

#Turning the 1s and 2s into M and F in the gender variable
clean_test$Gender <- ifelse(clean_test$Gender == "1", "M", "F")

#Turning the As and Bs into ADS and TD in the diagnosis variable using ifelse (short way)
clean_test$Diagnosis <- ifelse(clean_test$Diagnosis == "A", "ASD", "TD")

# Saving a csv. file named LanguageASD.csv to WD
write.csv(clean_test, file = "LanguageASD_test.csv")
```

```{r Loading new and csv file,  include=FALSE}
data = read.csv("LanguageASDv2.csv")
data_test = read.csv("LanguageASD_test.csv")
```


```{r Change names: training,  include=FALSE}
# Changing names in the training data
data <- rename.vars(data, "Diagnosis", "diagnosis")
data <- rename.vars(data, "Gender", "gender")
data <- rename.vars(data, "Age", "age")
data <- rename.vars(data, "Ethnicity", "ethnicity")
data <- rename.vars(data, "Visit", "visit")
data <- rename.vars(data, "ADOS1", "ados1")
data <- rename.vars(data, "MullenRaw1", "nonverbalIQ")
data <- rename.vars(data, "ExpressiveLangRaw1", "verbalIQ")

# Filtering out participants
data <- data[-c (1, 68, 69, 130, 131, 132, 276, 277, 284, 285), ]
```

```{r Change names: test,  include=FALSE}
data_test <- rename.vars(data_test, "Diagnosis", "diagnosis")
data_test <- rename.vars(data_test, "Gender", "gender")
data_test <- rename.vars(data_test, "Age", "age")
data_test <- rename.vars(data_test, "Ethnicity", "ethnicity")
data_test <- rename.vars(data_test, "Visit", "visit")
data_test <- rename.vars(data_test, "ADOS1", "ados1")
data_test <- rename.vars(data_test, "NonverbalIQ", "nonverbalIQ")
data_test <- rename.vars(data_test, "VerbalIQ", "verbalIQ")
```

### Exercise 1) Testing model performance

How did your models from last time perform? 
In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. 
Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 part 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

```{r Compare training and test data,  include=FALSE}
# data$CHI_MLU <- data$CHI_MLU[!is.na(data$CHI_MLU)]

# Removing NAs in data sets
data <- na.omit(data, cols = "CHI_MLU")
data_test <- na.omit(data_test, cols = "CHI_MLU")

# Using our best model from the last assignment
model_train<- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2) | ID), data = data, REML=FALSE)
model_test <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2) | ID), data = data_test, REML=FALSE)

summary(model_train)
# Root mean square error on the accuracy of the model's predictive power
rmse(data$CHI_MLU, predict(model_train, data))
# [1] 0.2775268
# Fairly low RMSE = okay predictive power

# RMSE on the test data
rmse(data_test$CHI_MLU,predict(model_train, data_test, allow.new.level = T))
# [1] 0.8739594
# Very bad prediction on the test set. High rmse = low accuracy
```

ANSWER:
We observe that the accuracy of the predictions decreases a lot when the model is applied to the new (test) data. The root mean squared error goes from 0.25 to 0.78, which indicates that the model’s predictive power has low accuracy. Where a root mean square error of 0.25 does not point towards a perfectly accurate model prediction of our training data, the RMSE of the test data is a lot less accurate. The model can thus be granted good explanatory power, but the predictive power is weak. 


### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model 
(Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)

Which model is better at predicting new data: 
1) The one you selected last week or
2) the one chosen via cross-validation this week?

- Test both of them on the test data.
- Report the results and comment on them.

```{r, include=FALSE}
# Creating the basic model
hercules <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = data, REML = F)

# Creating our model from last week
zeus <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2)| ID), data = data, REML=FALSE)
summary(zeus)
```

```{r Split: Cross validation, include=FALSE}
### BASIC MODEL - HERCULES
# Implement the 5-fold cross-fold plan
splitPlan <- kWayCrossValidation(nrow(data), 5, NULL, NULL)

# Examine the split plan
# str(splitPlan)

# Run the 5-fold cross validation plan from splitPlan on hercules model
k <- 5 # Number of folds

# Setting the variable for the predictive values from the CV
data$pred.cv <- 0 

# Creating loop
for(i in 1:k) {
  split <- splitPlan[[i]]
  model <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = data[split$train, ], REML = F)
  data$pred.cv[split$app] <- predict(model, newdata = data[split$app, ])
  }

# RMSE for basic model (hercules)
rmse(data$CHI_MLU, predict(hercules))
# [1] 0.3380725

# RMSE of the cross-validation predictions
rmse(data$CHI_MLU, data$pred.cv)
# [1] 0.5332898



### BEST MODEL - ZEUS
# Run the 5-fold cross validation plan from splitPlan on our best model (hercules)
data$pred.cv.best <- 0 

for(i in 1:k) {
  split1 <- splitPlan[[i]]
  model2 <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ados1 + MOT_MLU + verbalIQ + (1+visit + I(visit^2)+ I(visit^3)| ID), data =   data[split1$train, ], REML=FALSE)
  data$pred.cv.best[split1$app] <- predict(model2, newdata = data[split1$app, ])
  }

# Get the RMSE of the full model's predictions
rmse(data$CHI_MLU, predict(zeus))
## [1] 0.2527097

# Get the RMSE of the cross-validation predictions
rmse(data$CHI_MLU, data$pred.cv.best)
## [1] 0.4264432

```

How does the models predict the test data in reality?
```{r, Reality test, include=FALSE}
# Basic model
rmse(predict(hercules, data_test), data_test$CHI_MLU)
#[1] 1.068573

# Best model
rmse(predict(zeus, data_test), data_test$CHI_MLU)
#[1] 0.7793297
```

```{r Cross validation: createFolds, include=FALSE}
# Another way to do CV (Riccardo's way)
#k <- 10 #the number of folds
#Crate the folds
folds <- cvFolds(NROW(data), K=k)

# From class: Important to make the subject unique, as each subject, of coruse, is correlated with themselves
# createFolds(unique(data$SUBJECT))

# Set up column in dataframe
data$cv.pred <- rep(0,nrow(data))

# The loop
for(i in 1:k){
  train <- data[folds$subsets[folds$which != i], ] #Set the training set
  validation <- data[folds$subsets[folds$which == i], ] #Set the validation set
  newlm <- lmer(CHI_MLU ~ visit + diagnosis + (1 + visit | ID), data = train, REML = F) 
  #Get your new linear model (just fit on the train data)
  newpred <- predict(newlm,newdata=validation) 
  #Get the predicitons for the validation set (from the model just fit on the train data)
  data[folds$subsets[folds$which == i], ]$cv.pred <- newpred #Put the hold out prediction in the data set for later use
  }

# How does the basic model perform
rmse(data$cv.pred, data$CHI_MLU)
```

The "best" model from last assignment
```{r, include=FALSE}
#preparing a column
data$cv.pred_best <- rep(0,nrow(data))

for(i in 1:k){
  train <- data[folds$subsets[folds$which != i], ] #Set the training set
  validation <- data[folds$subsets[folds$which == i], ] #Set the validation set

  newlmer <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + (1+visit + I(visit^2) | ID), data = train, REML=FALSE)
  #Get your new linear model (just fit on the train data)
  
  newpred_best <- predict(newlmer, newdata=validation) 
  #Get the predicitons for the validation set (from the model just fit on the train data)

  data[folds$subsets[folds$which == i], ]$cv.pred_best <- newpred_best 
  #Put the hold out prediction in the data set for later use
  }

rmse(data$cv.pred_best, data$CHI_MLU)
```

ANSWER:
The basic linear model (pseudocode: mean length of utterance ~ visit + diagnosis + (1 + visit | ID)) produces a root mean squared error of only 0.34, when predicting on the training data set. However, when the model is trained with a 5-fold cross validation, the root mean squared error increases to 0.51. 

As already mentioned earlier the cubic model, that we chose to be the best model in the previous assignment (pseudocode: mean length of utterance ~ visit + visit^2 + visit^3 + verbalIQ + ados + mother’s MLU (1 + visit + visit^2 + visit^3| ID)) produces a mean squared error of 0.28, when applied to the traning data. As with the linear model, the error increases when the model is trained with a 5-fold cross validation - to 0.46, respectively. 

First of all, this means that the cubic model is still better at predicting the traning data compared to the linear model. 

When the two models’ cross validation predictions are applied to the actual test data they both turn out to be quite far off. The linear model produces a root mean squared error of 1.07. The cubic model produces a rmse of 0.79. Thus, the cubic model is still performing better on the test data than the linear, however what worse than predicted with the traning data (both with and without cross validation of the model). 



- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.
```{r, include=FALSE}
# Implement the 5-fold cross-fold plan. Reusing earlier splitPlan + k
data$pred.cv.create <- 0 

## Nannas model
for(i in 1:k) {
  split.create <- splitPlan[[i]]
  model.create <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ados1 + verbalIQ + MOT_MLU + types_CHI + (1+visit + I(visit^2)+   I(visit^3)| ID), data = data[split.create$train, ], REML=FALSE)
  data$pred.cv.create[split.create$app] <- predict(model.create, newdata = data[split.create$app, ])
  }
#            [1] 0.3737171
# Predicted: [1] 0.5446184

## P-hacking model
for(i in 1:k) {
  split.create <- splitPlan[[i]]
  model.create <- lmer(CHI_MLU ~ visit + I(visit^2) + I(visit^3) + ID + nonverbalIQ + verbalIQ + types_CHI + MOT_MLU + (1+visit + I(visit^2)+ I(visit^3)| ID), data = data[split.create$train, ], REML=FALSE)
  data$pred.cv.create[split.create$app] <- predict(model.create, newdata = data[split.create$app, ])
  }
#            [1] 0.3710135
# Predicted: [1] 0.5311729

## Quadratic model
for(i in 1:k) {
  split.create <- splitPlan[[i]]
  model.create <- lmer(CHI_MLU ~ visit + I(visit^2) + ados1 + MOT_MLU + (1+visit + I(visit^2) | ID), data = data[split.create$train, ], REML=FALSE)
  data$pred.cv.create[split.create$app] <- predict(model.create, newdata = data[split.create$app, ])
  }
# [1] 0.4516293

 

# RMSE on the cv model
rmse(data$pred.cv.create, data$CHI_MLU)
# RMSE on the predicted model
rmse(predict(model.create, data_test), data_test$CHI_MLU)
  

### We want to have our cake and eat it as well 
# It is a good idea to split in 5 parts (why 5 = )

#We want to make a loop For (F in Folds) select(ID in F)
#Take all the kids that a not in the F fold, and that i our train data set. 
#N=1 in the beginning, Rmse(train[n] =
#N = N+1 
```

Using Cross Validation to calculate accuracy of predictions and using the root mean squared error when applying the actual test data, the best predictive model is our p-hacked model from the previous assignment (pseudocode: Child MLU ~ visit + I(visit^2) + I(visit^3) + ID + nonverbalIQ + verbalIQ + types_CHI + MOT_MLU + (1 + visit + I(visit^2)+ I(visit^3)| ID). Interestingly, this model does not include ados or diagnosis.


### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.
```{r, include=FALSE}
# Find Bernie before the data was made anonymous
filter(test, ID == "Bernie")

# Filter out Bernie form the data_test dataframe
sanders <- filter(data_test, ados1 == "7", verbalIQ == "28")

```

You want to evaluate:
- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.
(Tip: recreate the equation of the model: Y=Intercept+BetaX1+BetaX2, etc; input the average of the TD group for each parameter in the model as X1, X2, etc.).
- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r How Bernie fares, include=FALSE}
detach(package:plyr)
library(dplyr)

summary(zeus)

# Filtering out TD participants
TD_average <- data %>% 
  filter(diagnosis == "TD")

# As numeric
TD_average$visit <- as.numeric(TD_average$visit)

# Taking the mean, grouping by visit and summarising to the mean MLU
TD_average <- TD_average %>% 
  group_by(visit) %>%
  summarise(mean_TD = mean(CHI_MLU, na.rm =T))

# Add column with Bernie's MLU for each visit
TD_average$bernie <- sanders$CHI_MLU

# Calculating the absolute difference between Bernie and the average TD child
TD_average$absolute_difference <- TD_average$bernie-TD_average$mean_TD

# Adding column with model predictions for Bernie
TD_average$prediction <- predict(zeus, sanders)

# Adding column calculating the absolute accuary of the model's predictions
TD_average$accuracy <- TD_average$prediction-TD_average$bernie
```

```{r, include=FALSE}
# Plotting Bernies predictions versus actual performance 
theme_set(theme_bw())
feelthebern <- ggplot(TD_average, aes(x = visit)) + 
  geom_smooth(aes(y=bernie), colour="red", method = "lm", formula = y ~ poly(x, 3), se = F) + 
  geom_smooth(aes(y=prediction), colour="blue", method = "lm", formula = y ~ poly(x, 3), se = F) +
  xlab('Number of Visit') +
  ylab("Bernie's Mean Length of Utterance") 
```

```{r Dataframe and Plot}
# Dataframe
TD_average

# Plot
feelthebern
```

Bernie, a child with ASD, fares generally quite well in his development of MLU compared to the average of TD participant. We have used the raw data to calculate the average of the TD group which was compared to Bernie’s average. He is consistently outperforming the TD average,
e.g. with his MLU of 1.98 and the TD average of 1.3 in the first visit. He is thus above the average of the TD participants, which is an interesting point as he has ASD. When looking into the model prediction of how Bernie should develop, the model overshoots his performance in the first five visits, but in the sixth his MLU is better than what the model predicts it to be (Bernie = 3.45, Model = 3.37). 
When using the model to calculate the predictions on Bernies data only we see that Bernie performed close to expectations at the first visit but in general below expectations at the first 5 visits. However, at visit 6 Bernie performed above expectations. The difference between the models predictions and how he fares is, however, fairly small. This development can also be seen in the plot below, where the blue line indicates the predictions obtained by the model and the red line is Bernie’s actual performance. 

